---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Supplemental packages must be loaded before we begin.

```{r, results=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
```

## Loading and preprocessing the data

The contents of the [*zipped personal activity dataset*](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) are read into a data frame called *activity*. The dataset is assumed to reside in the current working directory. The **date** field is then reformatted as POSIXct, and I examine the structure and contents of the data frame.

```{r, echo=TRUE}
activity <- read.csv(unzip("activity.zip"), stringsAsFactors=FALSE)
activity$date <- ymd(activity$date)

str(activity)
head(activity)
```

## What is mean total number of steps taken per day?

First, I calculate the total number of steps taken by the subject in each day, store the result in a data frame called *act_sum*, and print the first few lines. Note that some days contain only NA values.

```{r, echo=TRUE}
act_sum <- activity %>%
           group_by(date) %>%
           summarize(totalsteps=sum(steps))
act_sum
```

Then I generate a histogram showing how the sample is distributed by total number of steps taken per day. NA values are ignored.

```{r, echo=TRUE}
hist(act_sum$totalsteps,
     breaks=10,
     xlim=c(0,25000),
     ylim=c(0,25),
     main="",
     col="blue",
     xlab="Total Steps per Day",
     ylab="Number of Days"
    )
```

Finally, I calculate and store the mean and median number of steps taken per day, again ignoring the NA values (i.e., excluding them rather than treating them as zeroes).

```{r, echo=TRUE}
mean_steps = mean(act_sum$totalsteps, na.rm=TRUE)
median_steps = median(act_sum$totalsteps, na.rm=TRUE)
```

The mean daily number of steps is **`r formatC(mean_steps, format='f', big.mark=',', digits=2)`**  
The median daily number of steps is **`r formatC(median_steps, format='d', big.mark=',')`**

## What is the average daily activity pattern?

First, I calculate the average number of steps taken within each five-minute time interval across days in the sample.

```{r, echo=TRUE}
act_pat <- activity %>%
           group_by(interval) %>%
           summarize(meansteps=mean(steps,na.rm=TRUE))
head(act_pat)
```

Next, I generate a time-series plot showing how the average number of steps changes throughout the day.

```{r, echo=TRUE}
with(act_pat,
     plot(interval, meansteps,
          type="l",
          col=c("blue"),
          main="",
          xlab="Interval",
          ylab="Daily Average Steps"
         )
    )
```

Finally, I find the interval with the highest average daily steps.

```{r, echo=TRUE}
maxint <- act_pat[which(act_pat$meansteps==max(act_pat$meansteps)),]$interval
```

The interval labelled **`r maxint`** has the highest average step count.

## Imputing missing values

```{r, echo=TRUE}
nmiss <- nrow(activity[is.na(activity$steps),])
```

The *activity* dataset contains **`r formatC(nmiss, format='d', big.mark=',')`** observations with missing step counts.

I will replace missing values with the mean value for the corresponding interval and store the result in a dataset called *act_fix*.

```{r, echo=TRUE}
act_fix <- activity %>%
           left_join(act_pat, by="interval") %>%
           mutate(steps = ifelse(is.na(steps), meansteps, steps)) %>%
           select(-meansteps)
head(act_fix)
nmiss <- nrow(act_fix[is.na(act_fix$steps),])
```
The *act_fix* dataset contains **`r formatC(nmiss, format='d', big.mark=',')`** missing observations.

Now I generate a histogram showing how the sample is distributed by the total number of steps taken per day, after imputation.

```{r, echo=TRUE}
act_fix_sum <- act_fix %>%
               group_by(date) %>%
               summarize(totalsteps=sum(steps))
act_fix_sum
hist(act_fix_sum$totalsteps,
     breaks=10,
     xlim=c(0,25000),
     ylim=c(0,25),
     main="",
     col="red",
     xlab="Total Steps per Day",
     ylab="Number of Days"
    )
```

Finally, I calculate and store the mean and median number of steps taken per day.

```{r, echo=TRUE}
mean_fix_steps = mean(act_fix_sum$totalsteps)
median_fix_steps = median(act_fix_sum$totalsteps)
```

The mean daily number of steps after imputation is **`r formatC(mean_fix_steps, format='f', big.mark=',', digits=2)`**  
The median daily number of steps after imputation is **`r formatC(median_fix_steps, format='d', big.mark=',')`**

The mean and median remain unchanged as a result of my method of imputation. This is due to the fact that all missing values corresponded to days without any observations.  In effect, I added a handful of new data points at the mean, which had no effect on the result of either calculation.

## Are there differences in activity patterns between weekdays and weekends?

First I create a new factor variable called *daytype* which identifies date values as either weekday or weekend days.  I then calculate mean daily step counts for each interval and day type and store the result in a new *act_pat* dataset.

```{r, echo=TRUE}
activity$daytype <- factor(ifelse(weekdays(activity$date, abbreviate=TRUE) %in% c("Sat","Sun"), 1, 2), labels=c("weekend", "weekday"))
table(activity$daytype)
act_pat <- activity %>%
           group_by(daytype, interval) %>%
           summarize(meansteps=mean(steps,na.rm=TRUE))
head(act_pat)
```

Now I use the new variable to construct a panel plot enabling comparison of weekend and weekday average daily step patterns.

```{r, echo=TRUE}
ggplot(data=act_pat, aes(x=interval, y=meansteps)) +
  geom_line(color="green") +
  facet_grid(daytype~.) + 
  xlab("Interval") +
  ylab("Number of Steps") +
  theme_grey()
```

Certain differences are evident between weekend days and weekdays.  The peak at interval **835** appears in the weekday series but does not appear on weekends.  Weekend activity begins and ends later and tends to remain at a higher sustained level throughout the active portion of the day.